---
title: "Assignment 1"
output: html_document
---

```{r}


library(readr)

setwd("C:/Users/vajir/Documents/Broad Workshops/R-workspace/Assignment1/Assignment1-R-markdown/")

paad_rsem_norm_meta <- read_tsv("paad_rsem_norm_meta.txt")



```

How many rows and columns

```{r}

rows_columns <- dim(paad_rsem_norm_meta)

paste("Rows:", rows_columns[1])
paste("Columns:", rows_columns[2])

```

When you import a table, cleaning is important.
1 st step, subset data columns and non-data (meta) columns using grep() function

How many columns start with "TCGA"
Subset [] the colnames using the grep(), to find columns begin with TCGA, "^TCGA"

These objects (data.cols and meta.cols) are character vectors. They do not come as dataframes


```{r}

data.cols <- colnames (paad_rsem_norm_meta) [grep("^TCGA", colnames(paad_rsem_norm_meta))]

meta.cols <- colnames (paad_rsem_norm_meta) [grep("^TCGA", colnames(paad_rsem_norm_meta), invert = TRUE)]


paste(length(data.cols))

paste(meta.cols)

paste(class(data.cols))

paste(class(meta.cols))

```

How many columns are neither tumor (01), nor normal (11)? What are they?

These are the ones neither 01 or 11 of TCGA columans. These samples are metastatic samples - subset them as other.cols

To do this, first subset [] tumor and normal columns from the data.col object (a character object) using grepl().

grepl() returns a logical vector.

Next, use the inver (!) of value matching (%in%) between initially defined all data columns (data.cols) and tumor and normal [c(tumor.cols, normal.cols)], to subset the metastatic column.

data.cols is a character vector. grepl() returns two logical vectors. Value matching returns all TRUE objects from data.col. By inverting (!) we extract the answer

```{r}

tumor.cols <- data.cols[grepl ("TCGA-.*-.*-01.*-.*-.*-.",data.cols)]

normal.cols <- data.cols[grepl ("TCGA-.*-.*-11.*-.*-.*-.",data.cols)]

other.cols <- data.cols [! (data.cols %in% c(tumor.cols, normal.cols))]

other.cols

```

Here on keep working with the data frame

Generate a new dataset with only primary solid tumors. meta.cols are used as row names and tumor.cols are used as columns

```{r}

tumor.data <- paad_rsem_norm_meta [, c(meta.cols,tumor.cols)]

dim(tumor.data)

class(tumor.data)

colnames(tumor.data)



```


Remove any rows with unkonwn gene symbols or duplicated gene ids in e_gene_symbol column

create a new character vector by subsetting e_gene_symbol column of tumor.data data frame for duplicated gene ids and define dup.ids.

Then subset tumor.data data frame using NOT(!) operator, by value matching duplicated ids (dup.ids - the character vector) to e_gene_symbol column in tumor.data to define the dataframe tumor.data.v

```{r}

dup.ids <- tumor.data$e_gene_symbol [duplicated(tumor.data$e_gene_symbol)]

class(dup.ids)

paste(dup.ids)

tumor.data.v <- tumor.data [!tumor.data$e_gene_symbol %in% dup.ids,]

dim(tumor.data.v)

class(tumor.data.v)



```


Calculate the median across all samples

I intend to use rowMedian() from matrixStats package. 

So first I call the package by using library(), Then I got an error saying I only can use rowMedian function on either a matrix or vector.

Therefore, I converted the dataframe to a matrix, which eliminates the rownames by converting them to NA. Then I used the rowMedian(), with the operator na.rm=TRUE, so all NA will be removed from the calculation

```{r}

library("matrixStats",lib.loc = "~/R/win-library/3.4/")

tumor.data.v.matrix <- data.matrix(tumor.data.v,rownames.force = NA)

class(tumor.data.v.matrix)

row.median <- rowMedians(tumor.data.v.matrix, na.rm = TRUE)

class (row.median)

# assign rowmedian to original dataset using cbind. this will add the 49th column


tumor.data.v.median <- cbind(tumor.data.v, row.median)

colnames(tumor.data.v.median)


# Now, order according to the row median using order function, with operator, decreasing = TRUE 

tumor.data.v.ranked <- tumor.data.v.median[order(tumor.data.v.median$row.median, decreasing = TRUE),]

class(tumor.data.v.ranked)

dim(tumor.data.v.ranked)

colnames(tumor.data.v.ranked)

# Now subset top 250 rows

tumor.data.top.250 <- tumor.data.v.ranked [1:250,]

class(tumor.data.top.250)

dim(tumor.data.top.250)

colnames(tumor.data.top.250)


```


Generate a histogram for the distribution of expression across top 250 genes. Use log10(RSEM+1) for plotting


```{r}

# histograms use y axis for counts. Therefore, you cannot asign a variable for y axis. Only X axis could be defined.

library(ggplot2)

ggplot(tumor.data.top.250, aes ( x = log10(row.median+1))) + 
  
theme_bw() +
  
geom_histogram() + 
  
labs (X = "log 10 (RSEM+1)", title = " Distribution of Expression Across the Top 250 Genes")



ggplot(tumor.data.top.250, aes ( x = log10(row.median+1))) + 
  
theme_bw() +
  
geom_density() + 
  
labs (X = "log 10 (RSEM+1)", title = " Distribution of Expression Across the Top 250 Genes")


```



Only for primary solid tumors, generate a scatter plot for comapring TCGA-2J-AAB1-01A-11R-A41B-07 (X-axis) and TCGA-2J-AAB4-01A-12R-A41B-07 (y-axis) using log10(RSEM+1), add a regression line using lm()


```{r}

# First I will subset the data I want to plot

subset.tumor <- tumor.data.top.250[, c("TCGA-2J-AAB1-01A-11R-A41B-07","TCGA-2J-AAB4-01A-12R-A41B-07")]

dim(subset.tumor)

colnames(subset.tumor)

# The column name contains "-" followed by a capital letter which is misidentified as an object by R, therefore, I rename it. Is there a way to change all column names?

# Here you can also use gsub() in the begining to change all "-" to "_" in a given dataset

colnames(subset.tumor)[colnames(subset.tumor)=="TCGA-2J-AAB1-01A-11R-A41B-07"] <- "TCGA_2J_AAB1_01A_11R_A41B_07"

colnames(subset.tumor)[colnames(subset.tumor)=="TCGA-2J-AAB4-01A-12R-A41B-07"] <- "TCGA_2J_AAB4_01A_12R_A41B_07"

colnames(subset.tumor)


ggplot(subset.tumor, aes(x = TCGA_2J_AAB1_01A_11R_A41B_07, y = TCGA_2J_AAB4_01A_12R_A41B_07)) + geom_point() + geom_smooth(method=lm)



```














